Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-05-28 analysis/paper/paper.qmd
editing

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "Large cervidae", "Cervidae large size")) %>%
  mutate(taxa = str_replace(taxa, "Testudines", "Geoemydidae")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cervus") ~ "Sika deer", str_detect(taxa, "Bub") ~
                                     "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~
                                     "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~
                                     "Ray-finned fish", str_detect(taxa, "Geo") ~ "Turtle",
                                   str_detect(taxa, "Cervidae") ~ "Deer", taxa == "Sus scrofa domestica" ~
                                     "Domestic pig", taxa == "Aves" ~ "Bird")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "munt") ~ 3, str_detect(taxa, "Sus") ~ 4, 
                           str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 5,  str_detect(taxa, "Rat") ~ 6,
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Geo") ~ 8,  
                           str_detect(taxa, "Ave") ~ 9,  str_detect(taxa, "large m") ~ 10, 
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(head(., 13), across(where(is.numeric), sum), 
                      across(where(is.character), ~"Total determinate"))) %>%
  bind_rows(summarise(.[14:16, ], across(where(is.numeric), sum), 
                      across(where(is.character), ~"Total indeterminate"))) %>%
  bind_rows(summarise(head(., 16), across(where(is.numeric), sum), 
                      across(where(is.character), ~"Total"))) %>%
  #bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
  mutate(`Weight (g)` = round(`Weight (g)`, 1)) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|str_detect(`Common name`, "Total"), "-", `Common name`)) %>%
  mutate(Taxon = ifelse(Taxon == "Reeves's muntjac", "Muntiacus reevesi", Taxon)) %>%
  mutate(Taxon = ifelse(Taxon == "Cervus nippon", "Cervus nippon taiouanus", Taxon))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
deer_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")) %>% summarise(sum(NISP)) #|str_detect(Taxon, "Ree")
suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus")) %>% summarise(sum(NISP))
buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")) %>% summarise(sum(NISP))
deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 2)*100
suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100
buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
```

The faunal assemblage from the sampling area of Kiwulan consists of `r total_NISP` bone fragments, of which were identified to at least the family level (@tbl-counts). Among the mammal bones identified to the family level, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%). To further compare general patterns of animal consumption based on economic values, animals from the same family and of similar body size were grouped together. Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category. Similarly, domestic pigs (CL4: n = 2, CL6: n = 1) were integrated into the suid category. Rats (*Rattus sp.*) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).

>>>>

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "Large cervidae", "Cervidae large size")) %>%
  mutate(taxa = str_replace(taxa, "Testudines", "Geoemydidae")) %>%
  #mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          #TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cervus") ~ "Sika deer", str_detect(taxa, "Bub") ~
                                     "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~
                                     "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~
                                     "Ray-finned fish", str_detect(taxa, "Geo") ~ "Turtle",
                                   str_detect(taxa, "Cervidae") ~ "Deer", taxa == "Sus scrofa domestica" ~
                                     "Domestic pig", taxa == "Aves" ~ "Bird")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "munt") ~ 3, str_detect(taxa, "Sus") ~ 4, 
                           str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 5,  str_detect(taxa, "Rat") ~ 6,
                           str_detect(taxa, "Geo") ~ 7, str_detect(`Common name`, "fish") ~ 8,  
                           str_detect(taxa, "Ave") ~ 9,  str_detect(taxa, "Large m") ~ 10, 
                           str_detect(taxa, "Medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(head(., 16), across(where(is.numeric), sum), 
                      across(where(is.character), ~"Total"))) %>%
  #bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
  mutate(`Weight (g)` = round(`Weight (g)`, 1)) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|str_detect(`Common name`, "Total"), "-", `Common name`)) %>%
  mutate(Taxon = ifelse(Taxon == "Reeves's muntjac", "Muntiacus reevesi", Taxon)) %>%
  mutate(Taxon = ifelse(Taxon == "Cervus nippon", "Cervus nippon taiouanus", Taxon))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% slice(12:16) %>% summarise(sum(NISP))
family_ident <- total_NISP - unident
cervid_NISP <- fauna_combined_taxa %>% filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "ree")) %>% summarise(sum(NISP))
suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus")) %>% summarise(sum(NISP))
buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")|str_detect(Taxon, "Bos")) %>% summarise(sum(NISP))
cer_NISP_per <- round(cervid_NISP/(total_NISP-unident), 2)*100
suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100
buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
```

The faunal assemblage from the sampling area of Kiwulan consists of `r total_NISP` bone fragments, of which `r family_ident` were identified to at least the family level (@tbl-counts). Among those bones identified to the family or more specific level, Cervidae is the majority (`r cer_NISP_per`%), followed by Suidae (`r suid_NISP_per`%) and Bovinae (`r buff_NISP_per`%). To further compare general patterns of animal consumption based on economic values, animals from the same family and of similar body size were grouped together. Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category. Similarly, domestic pigs (CL4: n = 2, CL6: n = 1) were integrated into the suid category. Rats (*Rattus sp.*) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
